---
title: "NFL Football Penalty Visualization"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


The goal of this quick project is to visualize the penalties given to the San Francisco 49ers during the 2020 season. In NFL, each penalty pushes a team back from scoring opportunities. We will look at penalties received by Niners'offense and defense team.


First we load the necessary libraries for plotting.
```{r libraries, warning=FALSE, message=FALSE}
# load library
library(nflfastR)
library(tidyverse)
library(ggthemes)
library(forcats)

# Suppress summarize info
options(dplyr.summarise.inform = FALSE)

# To prevent R from printing numbers in scientif notation
options(scipen = 9999)
```

Second, we obtain data via *nflfastR* package. We will read the season data (R data) scraped by the authors of the package. 


```{r echo = TRUE, warning = FALSE, message=FALSE}
# Read data (2020 NFL season)
data <- readRDS(url('https://raw.githubusercontent.com/guga31bb/nflfastR-data/master/data/play_by_play_2020.rds'))

# Find San Francisco games only
sf = data %>% filter(home_team == 'SF' | away_team == 'SF')
```



Once we have the data for SF 49ers, we can visualize penalties received by both the offense and defense via horizontal bar plot.

First, let us visualize the penalties for all the violations.

```{r penalty, echo=TRUE, warning = FALSE, fig.height = 5, fig.width = 13}

penalties = sf %>% select(penalty_type,penalty_yards) %>% group_by(penalty_type) %>% filter(row_number() == n())%>% drop_na() %>% ungroup()

initial_plot <- penalties %>%
  # reorder penalty type by penalty yards
  mutate(penalty_type = fct_reorder(penalty_type, penalty_yards)) %>%
  # make a bar plot of penalty yards by penalty type
  ggplot(aes(x = penalty_type,y=penalty_yards)) + 
  geom_col()

titled_plot <- initial_plot + 
  # Add the title, subtitle, and caption
  labs(title = "How Severe is a Game Rule Violation?",
       subtitle = "In-Game Yard Penalties Received by the 49ers",
       #caption = "Source: nflfastR", 
       # Remove the x- and y-axis labels
       x = "", 
       y = "") 

flipped_plot <- titled_plot + 
  # Flip the axes
  coord_flip() + 
  # Remove the x-axis ticks and labels
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank())


#flipped_plot
flipped_plot + geom_text(aes(label = penalty_yards, 
                             y = penalty_yards + .03), 
                         position = position_dodge(0.9),
                         hjust = -.8,
                         vjust = .7)+ theme_fivethirtyeight()


```

For offense, the Niners were mostly caught for offensive holding and false start.

```{r offense, echo=TRUE, warning = FALSE, message = FALSE, fig.height = 5, fig.width = 13}
# visualize penalty counts

# Better way to pick up SF penalties on offense
foul = sf %>% filter(posteam == 'SF',penalty_team == 'SF')
foul = foul %>% select(penalty_type) %>% group_by(penalty_type) %>% summarize(count=n()) %>% drop_na() %>% ungroup()

initial_plot <- foul %>%
  # reorder penalty type by count
  mutate(penalty_type = fct_reorder(penalty_type, count)) %>%
  # make a bar plot of count by penalty type
  ggplot(aes(x = penalty_type,y=count)) + 
  geom_col()

titled_plot <- initial_plot + 
  # Add the title, subtitle, and caption
  labs(title = "Offense Penalties",
       subtitle = "San Francisco 49ers Season 2020",
       caption = "Source: nflfastR", 
       # Remove the x- and y-axis labels
       x = "", 
       y = "") 

flipped_plot <- titled_plot + 
  # Flip the axes
  coord_flip() + 
  # Remove the x-axis ticks and labels
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank())


#flipped_plot
flipped_plot + geom_text(aes(label = count, 
                             y = count + .03), 
                         position = position_dodge(0.9),
                         hjust = -.8,
                         vjust = 1)+ theme_fivethirtyeight()
```

For defense, the Niners were top for defensive holding and defensive pass interference.

```{r defense, echo=TRUE, warning = FALSE, message=FALSE, fig.height = 5, fig.width = 13}
# Better way to pick up SF penalties on defense
def_foul = sf %>% filter(defteam == 'SF',penalty_team == 'SF')
def_foul = def_foul %>% select(penalty_type,penalty) %>% group_by(penalty_type) %>% summarize(count=n()) %>% drop_na() %>% ungroup()

initial_plot <- def_foul %>%
  # reorder penalty type by count
  mutate(penalty_type = fct_reorder(penalty_type, count)) %>%
  # make a bar plot of count by penalty type
  ggplot(aes(x = penalty_type,y=count)) + 
  geom_col()

titled_plot <- initial_plot + 
  # Add the title, subtitle, and caption
  labs(title = "Defense Penalties",
       subtitle = "San Francisco 49ers Season 2020",
       caption = "Source: nflfastR", 
       # Remove the x- and y-axis labels
       x = "", 
       y = "") 

flipped_plot <- titled_plot + 
  # Flip the axes
  coord_flip() + 
  # Remove the x-axis ticks and labels
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank())
#flipped_plot


flipped_plot + geom_text(aes(label = count, 
                             y = count + .03), 
                         position = position_dodge(0.9),
                         hjust = -1,
                         vjust = 1)+ theme_fivethirtyeight()

```